### 2014-10-19 PAPER_AMMLY.tex
Making space for Experiments

To the best of our knowledge the recent work of~\cite{DBLP:journals/corr/DauphinPGCGB14} is the first one to explore the statistical properties of neural network error surfaces through the prism of statistical physics and random matrix theory. In particular these theoretical works~\cite{Bray2007,Fyodorov2007,Baldi:1989:NNP:70359.70362,wigner_semicircle} suggest the existence of a certain structure of critical points of random Gaussian error functions on high dimensional continuous spaces. They imply that critical points which error is much higher than the global minimum are exponentially likely to be saddle points with many negative and approximate plateau directions whereas all local minima are likely to have an error very close to that of the global minimum (these results are conveniently reviewed in~\cite{DBLP:journals/corr/DauphinPGCGB14}). The work of~\cite{DBLP:journals/corr/DauphinPGCGB14} establishes a strong connection between neural networks and the theory of random Gaussian fields through an empirical evaluation by providing an experimental evidence that the cost function of neural networks exhibits the same properties as the Gaussian error functions on high dimensional continuous spaces. Nevertheless they provide no theortical justification for the existence of this connection which instead we provide in this paper.

@@

\begin{figure*}[htp!]
  \center
\includegraphics[width = 2.5in]{Theta_cp.pdf} 
\includegraphics[width = 2.5in]{Theta_lm_sp.pdf}
\caption{Functions $\Theta_H(u)$ and $\Theta_{k,H}(u)$ for $k = \{0,1,\dots,5\}$. Parameter $H$ was set to $H = 3$. Black line: $u = -E_0(H)$, red line: $u = -E_{\infty}(H)$.}
\label{fig:Thetas}
\end{figure*}

@@

Let us now introduce the number that we will refer to as $E_{\infty}$ and use later on.
\begin{equation}
E_{\infty} = E_{\infty}(H) = 2\sqrt{\frac{H-1}{H}}.
\label{eq:E_inf}
\end{equation}
We will also define two non-decreasing, continuous functions on $\mathbb{R}$, $\Theta_H$ and $\Theta_{k,H}$ as follows
\[\Theta_H(u) = \left \{
  \begin{tabular}{c}
  $\!\!\!\!\frac{1}{2}\log(H\!-\!1) \!-\! \frac{H-2}{4(H-1)} \!-\! I(u)$ $\:\:\:$if$\:\:$$u \!\leq\! -E_{\infty}$\\
  $\!\!\!\!\!\frac{1}{2}\log(H\!-\!1) \!-\! \frac{H-2}{4(H-1)}$ $\:\:\:\:\:\:$if$\:\:$$-E_{\infty} \!\leq\! u \leq 0$\\
  $\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\frac{1}{2}\log(H-1)$ $\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:$if$\:\:$$0 \!\leq\! u$
  \end{tabular}
\right.,
\]
and for any integer $k \geq 0$:
\[\Theta_{k,H}(u) \!=\! \left \{
  \begin{tabular}{c}
  $\!\!\!\!\!\frac{1}{2}\!\log(\!H\!-\!1\!) \!-\! \frac{H\!-\!2}{4(\!H-1\!)} \!-\! (k\!+\!1)I(u)$ if$\:$$u \!\!\leq\!\! -E_{\infty}$\\
  $\!\!\!\!\frac{1}{2}\log(\!H\!-\!1\!) \!-\! \frac{H\!-\!2}{4(\!H-1\!)}$ $\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:$if$\:$$u \!\!\geq\!\! -E_{\infty}$
  \end{tabular}
\right.
\]
where 
\[I(u) = -\frac{u}{E_{\infty}^2}\sqrt{u^2 - E_{\infty}^2} - \log(-u + \sqrt{u^2 - E_{\infty}^2}) + \log E_{\infty}.
\]
Note that the maximal values of $\Theta_H(u)$ and $\Theta_{k,H}(u)$ are respectively $\frac{1}{2}\log(H-1)$ and $\frac{1}{2}\log(H-1) - \frac{H-2}{H}$. Figure~\ref{fig:Thetas} captures exemplary plots of the functions $\Theta_H(u)$ and $\Theta_{k,H}(u)$ for $k = 0$. Also note that the following corollary holds.
\begin{corollary}
For all $k > 0$ and $u < -E_{\infty}$, $\Theta_{k,H}(u) < \Theta_{0,H}(u)$.
\label{cor:Theta}
\end{corollary}
Finally, for any integer $k \geq 0$, let $E_k = E_k(H) > 0 $ be defined as the unique solution to 
\[\Theta_{k,H}(-E_k) = 0.
\]
For any fixed $H \geq 3$, the sequence $\{E_k(H)\}_{k\in\mathbb{\Lambda}}$ is strictly decreasing and converges to $E_{\infty}$ as $k \rightarrow \infty$~\cite{AAC2010}. We will now define the ground state energy.

\begin{figure*}[htp!]
  \center
\includegraphics[width = 2.5in]{Distr_cp.pdf} 
\includegraphics[width = 2.5in]{Distr_lm.pdf}
\caption{Distribution of the mean number of critical points and local minima. Parameters $H$ and $\Lambda$ were set to $H = 3$ and $\Lambda = 1000$. Black line: $u = -E_0(H)$, red line: $u = -E_{\infty}(H)$, blue line: $u = 0$.}
\label{fig:Distr_cp_lm}
\end{figure*}

\begin{definition}
The ground state energy is the (normalized) minimum of the loss function $\mathcal{L}_{\Psi,c,H}$
\[G^\Psi = \frac{1}{\Lambda}\inf_{{\bm w}\in \mathcal{S}_{\Psi,c}}\mathcal{L}_{\Psi,c,H}({\bm w}).
\]
\end{definition}
Next we provide theorems (resp. Theorem~\ref{thm:lb},~\ref{thm:ub},~\ref{thm:layer}, ~\ref{thm:lmsp} and~\ref{thm:cp}) which are direct consequences of the theoretical findings in~\cite{AAC2010} (resp. Theorem 2.12, 2.14, 2.15, 2.5 and 2.8)  and the boundeness of $c$ and $H$.  
\begin{theorem}
For every $H \geq 3$
\[\liminf_{\Lambda \rightarrow \infty} G^\Psi \geq -E_0(H),
\]
and for $H \geq 4$
\[\liminf_{\Lambda \rightarrow \infty} G^\Psi = -E_0(H) \:\:\:\text{in probability.}
\]
\label{thm:lb}
\end{theorem}
By Theorem~\ref{thm:lb}, for large-size networks it is improbable to find a critical value below level $-\Lambda E_0(H)$.
\begin{theorem}
For an integer $k \geq 0$ and $\epsilon > 0$,
\[\limsup_{\Lambda \rightarrow \infty} \frac{1}{\Lambda^2}\log(\mathbb{P}(\mathcal{C}_{\Lambda,k}(-E_{\infty}(H) + \epsilon,\infty)>0)) < 0
\]
\label{thm:ub}
\end{theorem}
Theorem~\ref{thm:ub} implies that for large-size networks all critical values of the loss function that are of non-diverging index must lie below the threshold $-\Lambda E_{\infty}(H)$. We will refer to this important threshold as an \textit{energy barrier}. Any critical point that lies above this threshold is a high-index saddle point with overwhelming probablity. Theorem~\ref{thm:lb} and~\ref{thm:ub} both imply that for large-size networks all critical values of the loss function that are of non-diverging index must lie in the band $\left(-\Lambda E_0(H),-\Lambda E_{\infty}(H)\right)$.
\begin{theorem}
For $k \geq 0$ and $\epsilon > 0$,
\[\limsup_{\Lambda \rightarrow \infty} \frac{1}{\Lambda^2}\log(\mathbb{P}(\sum_{i=k}^{\infty}\!\mathcal{C}_{\Lambda,i}(-E_k(H) - \epsilon,\infty) \!>\! 0)) \!<\! 0
\]
\label{thm:layer}
\end{theorem}
Theorem~\ref{thm:layer} implies that for large-size networks finding a critical value with index larger or equal to $k$ (for any fixed integer $k$) below energy level $-\Lambda E_k(H)$ is improbable.

@@

An interesting implication of Theorems~\ref{thm:lb},~\ref{thm:ub} and~\ref{thm:layer} is that recovering the global minimizer when starting from any (local) minima from one of the energy bands above, e.g. $\left(-\Lambda E_i(H),-\Lambda E_{i+1}(H)\right)$, diverges with $\Psi$ since it is bounded below by $\Lambda (E_0(H) - E_i(H))$.

Next we will show the logarithmic asymptotics of the mean number of critical points (the asymptotics of the mean number of critical points can be found in the Supplementary material).

\begin{figure*}[htp!]
  \center
\includegraphics[width = 2.2in]{Distr_lm_sp_li.pdf} 
\includegraphics[width = 2.2in]{Distr_lm_sp_li_zoomed_left.pdf} 
\includegraphics[width = 2.2in]{Distr_lm_sp_li_zoomed_right.pdf} 
\caption{Distribution of the mean number of local minima and low-index saddle points (oryginal and zoomed). Parameters $H$ and $\Lambda$ were set to $H = 3$ and $\Lambda = 1000$. Black line: $u = -E_0(H)$, red line: $u = -E_{\infty}(H)$.}
\label{fig:Distr_lm_sp}
\end{figure*}

\begin{theorem}
For all $H \geq 2$ and $k \geq 0$ fixed
\[\lim_{\Lambda \rightarrow \infty}\frac{1}{\Lambda}\log\mathbb{E}[\mathcal{C}_{\Lambda,k}(u)] = \Theta_{k,H}(u).
\]
\label{thm:lmsp}
\end{theorem}
\begin{theorem}
For all $H \geq 2$
\[\lim_{\Lambda \rightarrow \infty}\frac{1}{\Lambda}\log\mathbb{E}[\mathcal{C}_{\Lambda}(u)] = \Theta_{H}(u).
\]
\label{thm:cp}
\end{theorem}
Figure~\ref{fig:Distr_cp_lm} captures exemplary plots of the distributions of the mean number of critical points and local minima and Figure~\ref{fig:Distr_lm_sp} captures exemplary plots of the distributions of the mean number of local minima and low-index saddle points. Both figures clearly show that the local minima and low-index saddle points are located in the band $\left(-\Lambda E_0(H),-\Lambda E_{\infty}(H)\right)$ whereas high-index saddle points can only be found above the energy barrier $-\Lambda E_{\infty}(H)$. Figure~\ref{fig:Distr_lm_sp} also clearly reveals the layered structure for the lowest critical values of the loss function as predicted by the theory.

Note that from Theorem~\ref{thm:lmsp} and Corollary~\ref{cor:Theta} it follows that the number of critical points in the band $\left(-\Lambda E_0(H),-\Lambda E_{\infty}(H)\right)$ increases exponentially as $\Psi$ grows and that local minima dominate over saddle points and this domination also grows exponentially as $\Psi$ grows. Thus for large-size networks (large $\Psi$) the probability of recovering a saddle point in the band $\left(-\Lambda E_0(H),-\Lambda E_{\infty}(H)\right)$, rather than a local minima, goes to $0$.

@@

section{Conclusion and Future Work}
\label{sec:ConandFutWork}

@@

\newpage

@@

>>>>

To the best of our knowledge the recent work of~\cite{DBLP:journals/corr/DauphinPGCGB14} is the first one to explore the statistical properties of neural network error surfaces through the prism of statistical physics and random matrix theory. In particular these theoretical works~\cite{Bray2007,Fyodorov2007,Baldi:1989:NNP:70359.70362,wigner_semicircle} suggest the existence of a certain structure of critical points of random Gaussian error functions on high dimensional continuous spaces. They imply that critical points which error is much higher than the global minimum are exponentially likely to be saddle points with many negative and approximate plateau directions whereas all local minima are likely to have an error very close to that of the global minimum (these results are conveniently reviewed in~\cite{DBLP:journals/corr/DauphinPGCGB14}). The work of~\cite{DBLP:journals/corr/DauphinPGCGB14} establishes a strong connection between neural networks and the theory of random Gaussian fields through an empirical evaluation by providing an experimental evidence that the cost function of neural networks exhibits the same properties as the Gaussian error functions on high dimensional continuous spaces. Nevertheless they provide no theortical justification for the existence of this connection which instead we provide in this paper. \textcolor{red}{We should comment why saddle-point problem in not a problem for us.}

@@

@@

Let us now introduce the number that we will refer to as $E_{\infty}$ and use later on.
\begin{equation}
E_{\infty} = E_{\infty}(H) = 2\sqrt{\frac{H-1}{H}}.
\label{eq:E_inf}
\end{equation}
We will also define two non-decreasing, continuous functions on $\mathbb{R}$, $\Theta_H$ and $\Theta_{k,H}$ as follows
\[\Theta_H(u) = \left \{
  \begin{tabular}{c}
  $\!\!\!\!\frac{1}{2}\log(H\!-\!1) \!-\! \frac{H-2}{4(H-1)} \!-\! I(u)$ $\:\:\:$if$\:\:$$u \!\leq\! -E_{\infty}$\\
  $\!\!\!\!\!\frac{1}{2}\log(H\!-\!1) \!-\! \frac{H-2}{4(H-1)}$ $\:\:\:\:\:\:$if$\:\:$$-E_{\infty} \!\leq\! u \leq 0$\\
  $\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\frac{1}{2}\log(H-1)$ $\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:$if$\:\:$$0 \!\leq\! u$
  \end{tabular}
\right.,
\]
and for any integer $k \geq 0$:
\[\Theta_{k,H}(u) \!=\! \left \{
  \begin{tabular}{c}
  $\!\!\!\!\!\frac{1}{2}\!\log(\!H\!-\!1\!) \!-\! \frac{H\!-\!2}{4(\!H-1\!)} \!-\! (k\!+\!1)I(u)$ if$\:$$u \!\!\leq\!\! -E_{\infty}$\\
  $\!\!\!\!\frac{1}{2}\log(\!H\!-\!1\!) \!-\! \frac{H\!-\!2}{4(\!H-1\!)}$ $\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:\:$if$\:$$u \!\!\geq\!\! -E_{\infty}$
  \end{tabular}
\right.
\]
where 
\[I(u) = -\frac{u}{E_{\infty}^2}\sqrt{u^2 - E_{\infty}^2} - \log(-u + \sqrt{u^2 - E_{\infty}^2}) + \log E_{\infty}.
\]
Note that the maximal values of $\Theta_H(u)$ and $\Theta_{k,H}(u)$ are respectively $\frac{1}{2}\log(H-1)$ and $\frac{1}{2}\log(H-1) - \frac{H-2}{H}$. Figure~\ref{fig:Thetas} captures exemplary plots of the functions $\Theta_H(u)$ and $\Theta_{k,H}(u)$ for $k = 0$. Also note that the following corollary holds.

\begin{figure}[h]
  \center
\includegraphics[width = 1.6in]{Theta_cp.pdf}
\includegraphics[width = 1.6in]{Theta_lm_sp.pdf}
\vspace{-0.3in}
\caption{Functions $\Theta_H(u)$ and $\Theta_{k,H}(u)$ for $k = \{0,1,\dots,5\}$. Parameter $H$ was set to $H = 3$. Black line: $u = -E_0(H)$, red line: $u = -E_{\infty}(H)$. Figure must be read in color.}
\label{fig:Thetas}
%\vspace{-0.05in}
\end{figure}

\begin{corollary}
For all $k > 0$ and $u < -E_{\infty}$, $\Theta_{k,H}(u) < \Theta_{0,H}(u)$.
\label{cor:Theta}
\end{corollary}
Finally, for any integer $k \geq 0$, let $E_k = E_k(H) > 0 $ be defined as the unique solution to 
\[\Theta_{k,H}(-E_k) = 0.
\]
For any fixed $H \geq 3$, the sequence $\{E_k(H)\}_{k\in\mathbb{\Lambda}}$ is strictly decreasing and converges to $E_{\infty}$ as $k \rightarrow \infty$~\cite{AAC2010}. We will now define the ground state energy.

\begin{definition}
The ground state energy is the (normalized) minimum of the loss function $\mathcal{L}_{\Psi,c,H}$
\[G^\Psi = \frac{1}{\Lambda}\inf_{{\bm w}\in \mathcal{S}_{\Psi,c}}\mathcal{L}_{\Psi,c,H}({\bm w}).
\]
\end{definition}
Next we provide theorems (resp. Theorem~\ref{thm:lb},~\ref{thm:ub},~\ref{thm:layer}, ~\ref{thm:lmsp} and~\ref{thm:cp}) which are direct consequences of the theoretical findings in~\cite{AAC2010} (resp. Theorem 2.12, 2.14, 2.15, 2.5 and 2.8)  and the boundeness of $c$ and $H$.

\begin{figure*}[htp!]
  \center
\includegraphics[width = 1.65in]{Distr_cp.pdf}
\includegraphics[width = 1.65in]{Distr_lm_sp_li.pdf} 
\includegraphics[width = 1.65in]{Distr_lm_sp_li_zoomed_left.pdf} 
\includegraphics[width = 1.65in]{Distr_lm_sp_li_zoomed_right.pdf} 
\vspace{-0.3in}
\caption{Distribution of the mean number of critical points, local minima and low-index saddle points (oryginal and zoomed). Parameters $H$ and $\Lambda$ were set to $H = 3$ and $\Lambda = 1000$. Black line: $u = -E_0(H)$, red line: $u = -E_{\infty}(H)$. Figure must be read in color.}
\label{fig:Distr_cp_lm_sp}
\vspace{-0.1in}
\end{figure*}

\begin{theorem}
For every $H \geq 3$
\[\liminf_{\Lambda \rightarrow \infty} G^\Psi \geq -E_0(H),
\]
and for $H \geq 4$
\[\liminf_{\Lambda \rightarrow \infty} G^\Psi = -E_0(H) \:\:\:\text{in probability.}
\]
\label{thm:lb}
\end{theorem}
By Theorem~\ref{thm:lb}, for large-size networks it is improbable to find a critical value below level $-\Lambda E_0(H)$.
\begin{theorem}
For an integer $k \geq 0$ and $\epsilon > 0$,
\[\limsup_{\Lambda \rightarrow \infty} \frac{1}{\Lambda^2}\log(\mathbb{P}(\mathcal{C}_{\Lambda,k}(-E_{\infty}(H) + \epsilon,\infty)>0)) < 0
\]
\label{thm:ub}
\end{theorem}
Theorem~\ref{thm:ub} implies that for large-size networks all critical values of the loss function that are of non-diverging index must lie below the threshold $-\Lambda E_{\infty}(H)$. We will refer to this important threshold as an \textit{energy barrier}. Any critical point that lies above this threshold is a high-index saddle point with overwhelming probablity. Theorem~\ref{thm:lb} and~\ref{thm:ub} both imply that for large-size networks all critical values of the loss function that are of non-diverging index must lie in the band $\left(-\Lambda E_0(H),-\Lambda E_{\infty}(H)\right)$.
\begin{theorem}
For $k \geq 0$ and $\epsilon > 0$,
\[\limsup_{\Lambda \rightarrow \infty} \frac{1}{\Lambda^2}\log(\mathbb{P}(\sum_{i=k}^{\infty}\!\mathcal{C}_{\Lambda,i}(-E_k(H) - \epsilon,\infty) \!>\! 0)) \!<\! 0
\]
\label{thm:layer}
\end{theorem}
Theorem~\ref{thm:layer} implies that for large-size networks finding a critical value with index larger or equal to $k$ (for any fixed integer $k$) below energy level $-\Lambda E_k(H)$ is improbable.

@@

An interesting implication of Theorems~\ref{thm:lb},~\ref{thm:ub} and~\ref{thm:layer} is that recovering the global minimizer when starting from any (local) minima from one of the energy bands above, e.g. $\left(-\Lambda E_i(H),-\Lambda E_{i+1}(H)\right)$, diverges with $\Lambda$ since it is bounded below by $\Lambda (E_0(H) - E_i(H))$.

Next we will show the logarithmic asymptotics of the mean number of critical points (the asymptotics of the mean number of critical points can be found in the Supplementary material).
\begin{theorem}
For all $H \geq 2$ and $k \geq 0$ fixed
\[\lim_{\Lambda \rightarrow \infty}\frac{1}{\Lambda}\log\mathbb{E}[\mathcal{C}_{\Lambda,k}(u)] = \Theta_{k,H}(u).
\]
\label{thm:lmsp}
\end{theorem}
\begin{theorem}
For all $H \geq 2$
\[\lim_{\Lambda \rightarrow \infty}\frac{1}{\Lambda}\log\mathbb{E}[\mathcal{C}_{\Lambda}(u)] = \Theta_{H}(u).
\]
\label{thm:cp}
\end{theorem}
Note that from Theorem~\ref{thm:lmsp} and Corollary~\ref{cor:Theta} it follows that the number of critical points in the band $\left(-\Lambda E_0(H),-\Lambda E_{\infty}(H)\right)$ increases exponentially as $\Psi$ grows and that local minima dominate over saddle points and this domination also grows exponentially as $\Psi$ grows. Thus for large-size networks (large $\Psi$) the probability of recovering a saddle point in the band $\left(-\Lambda E_0(H),-\Lambda E_{\infty}(H)\right)$, rather than a local minima, goes to $0$.

Figure~\ref{fig:Distr_cp_lm_sp} captures exemplary plots of the distributions of the mean number of critical points, local minima and low-index saddle points. Clearly local minima and low-index saddle points are located in the band $\left(-\Lambda E_0(H),-\Lambda E_{\infty}(H)\right)$ whereas high-index saddle points can only be found above the energy barrier $-\Lambda E_{\infty}(H)$. Figure~\ref{fig:Distr_cp_lm_sp} also reveals the layered structure for the lowest critical values of the loss function.

@@

\section{Conclusion and Future Work}
\label{sec:ConandFutWork}

@@

\clearpage

@@

\section{Asymptotics of the mean number of critical points and local minima}
